---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# PacemakeR

<!-- badges: start -->
<!-- badges: end -->

PacemakeR is an R package for analysing pacing strategies in major marathon
races, such as London or Berlin. It provides simple tools to analyse marathon
split data, compute paces, summarise how runners distribute their effort over
the race, and quantify the advantage of one pacing strategy over another. It is
built as a proper R package — `roxygen2` documentation, a generated `NAMESPACE`,
declared dependencies, a bundled dataset, and a `testthat` suite — and developed
under version control on [GitHub](https://github.com/Runandecon/PacemakeR), from
which it installs directly.

## Introduction

In marathon running, a *negative split* means covering the second half of the
race faster than the first. It is widely regarded as the hallmark of mastery,
showcasing discipline and precision, and the latest world records have been set
this way — reflecting the delicate balance between conserving energy early and
maximising output late. For any marathon-running economist (like me), this
naturally becomes an optimization problem: start too slowly and you under-use
your physiological budget; start too fast and you collapse under nonlinear
fatigue constraints (the most common outcome, known as "hitting the wall").

Whether negative splitting or even splitting should be the goal is a matter of
debate. The negative-split approach is essentially risk minimization: it lets
you start conservatively and minimise blow-up risk, assuming you can set a
realistic goal. On a good day you negative split and reach the target; on a bad
day you miss the dream target but still even split. Since marathon performance
comes down to months and years of training, chance plays only a minor role, and
there is rarely a "surprise" performance outside a fairly predictable range.

This package analyses marathon split data through that lens. It evaluates a
runner's pacing relative to the rest of the field, which filters out external
factors such as weather, temperature, or course features like a hill at
kilometre 18. This comparative approach gives a cleaner read on pacing and lets
us describe the strategy the data actually rewards.

### A few running terms, briefly

For readers who have never run a marathon, three terms are worth defining up
front, since everything below is built on them:

- **Distance.** A marathon is 42.195 km.
- **Splits** are recorded by races as the the elapsed
  time between fixed checkpoints (timing mats) along the course, e.g. at 5 km,
  10 km, the half (21.1 km), and so on.
- **Pace.** Speed expressed as time per kilometre, written `mm:ss` (e.g. a pace
  of `04:30` means four minutes thirty seconds per km). Lower pace = faster.
  Pace is just the inverse of speed, scaled to the units runners actually think in.

## Installation

You can install the development version of PacemakeR from
[GitHub](https://github.com/Runandecon/PacemakeR) with:

``` r
# install.packages("devtools")
devtools::install_github("Runandecon/PacemakeR")
library(PacemakeR)
```

## The data

The package ships with `London_Marathon_2026`, the official split-level results
of the fastest 6000 amateur men in the 2026 London Marathon, scraped from the
official results site. Each row is one runner at one timing mat — the standard
way race results are published — which makes it *long-format* data: many rows
per runner.

Marathon results in the wild are messy: missing splits, runners who dropped out,
times recorded as blanks or `DNF`. Before any analysis is possible the data has
to be reshaped from this long format into one tidy pacing table per runner and
cleaned of those invalid entries. The package does this with a **tidyverse**
(`dplyr` / `tibble`) pipeline grouping by runner, ordering splits, and
differencing successive times into segment paces. This is the data-manipulation
backbone everything downstream relies on.

```{r download, echo=FALSE, message=FALSE, warning=FALSE}
library(devtools)
devtools::install_github("Runandecon/PacemakeR")
library(PacemakeR)
```

```{r data}
data("London_Marathon_2026")
head(London_Marathon_2026)
```

The columns are `Name`, `Nationality`, `Bib` (the runner's id number), `Distance` (the split
distance in km), and `Time` (elapsed time at that split, as `hh:mm:ss`).

## The functions

The functions build on each other: describe the field, distil its pacing
strategy, and price the payoff of running it well. Each one also leans on a
particular tool from the course, introduced where it does its work.

### `marathon_summary()` — describe the field

`marathon_summary()` reduces the field to a five-number summary (mean, median,
min, max, standard deviation) of either finishing times or any intermediate
split. It can report this in **time** (`hh:mm:ss`) or in **pace** (`mm:ss` per
km), and can draw the distribution.

For an econometrician: this is the descriptive-statistics step — what does the
distribution of outcomes look like, and where would a given target sit within
it? The distribution plot is the package's first use of **ggplot2**, which
handles all visualisation here. On top of the histogram it overlays vertical
reference markers; the two below mark a 2:30 and a 3:00 finish (elite-amateur
and strong-amateur benchmarks), which the function converts automatically from a
finish time into the pace it implies, so a target on the time scale lines up
correctly on the pace axis.

```{r summary-table}
marathon_summary(London_Marathon_2026)
```

```{r pacing-distribution}
marathon_summary(London_Marathon_2026, distance = 5, pace = TRUE, plot = TRUE, markers = c("02:30:00", "03:00:00"))

marathon_summary(London_Marathon_2026, plot = TRUE, markers = c("02:30:00", "03:00:00"))
```

### `pacemaker()` — distil the strategy

`pacemaker()` is the analytical core. For every runner it builds a
*relative-pace curve*: their pace at each split, normalised so that a chosen
reference split equals 1. Normalising this way is what makes a fast and a slow
runner comparable on the same axis — it strips out absolute speed and leaves
only the *shape* of the effort, i.e. where each runner spent or conserved
energy. The function then aggregates these curves across the whole field, with a
quantile band for the spread, and separately averages the runners who ran a
negative split, overlaying them for comparison.

For the econometrician: each runner's curve is an index normalised to a base
period (the reference split), exactly like a price index normalised to a base
year. Values above 1 mean "faster than the reference," below 1 "slower." The
aggregate curve is the field's average index path; the band is its empirical
spread.

This is where the package's use of **S3 classes** comes in. Rather than handing
back a bare list of numbers, `pacemaker()` returns an object of a custom class,
`pacemaker`, that bundles the curve, the cumulative-advantage data, and metadata
about the fit. Defining a class lets the package give the object its own `print`
and `plot` methods, so it behaves like a first-class R object: typing its name
dispatches to `pacemaker_print`, which lays out the pacing progression as a
table, and calling `plot()` dispatches to `pacemaker_plot`, which draws the
curve. The single object is the whole analysis — everything you might want to
see is computed once and read off it on demand.

```{r curve}
london <- pacemaker(London_Marathon_2026, relative = 21.0975)
```

```{r curve-plot}
pacemaker_print(london)
pacemaker_plot(london)
```

### `plot(opt, "gain")` — price the payoff

Because the analysis lives in one `pacemaker` object, pricing the strategy is
just another view of it rather than a separate function. The object also stores
the *gain*: the cumulative advantage a negative-split runner builds over the
field across the race. Asking its `plot` method for the `"gain"` view integrates
the gap between the two curves along the course and reports it either as distance
(metres ahead) or, given a reference pace, as time saved — the same S3 `plot`
method, switched to a different output.

For the econometrician: this is the cumulative integral of the per-segment
difference between two index paths — the running total of how far ahead the
negative-split strategy puts you, expressed in units a runner cares about.

```{r gain}
pacemaker_plot(london, "gain")                                   # advantage in metres
```

```{r gain-time}
pacemaker_plot(london, "gain", unit = "time", pace_sec = 300)    # same advantage, in seconds
```

## Course tools at a glance

The tools above were each introduced where they do their work: version control
and the package skeleton frame the whole project; the **tidyverse** pipeline
reshapes and cleans the raw splits; **ggplot2** carries every visualisation; and
a custom **S3 class** turns the central `pacemaker()` result into an object that
prints and plots itself. The package is tracked on GitHub, documented with
`roxygen2`, and validated by a `testthat` suite covering the helpers, the core
computation, and the S3 methods, so `devtools::check()` passes from a clean
environment.

## Notes on reproducibility

This `README.md` is generated from `README.Rmd`; re-render it with
`devtools::build_readme()` whenever the source changes, and commit the resulting
figures in `man/figures/` so they display on GitHub.
