---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# PacemakeR

<!-- badges: start -->
<!-- badges: end -->

PacemakeR is an R package for analysing pacing strategies in major marathon
races, such as London or Berlin. It provides simple tools to analyse marathon
split data, compute paces, summarise how runners distribute their effort over
the race, and quantify the advantage of ones pacing strategy over another. It is
built as a proper R package — `roxygen2` documentation, a generated `NAMESPACE`,
declared dependencies, a bundled dataset, and a `testthat` suite — and developed
under version control on [GitHub](https://github.com/Runandecon/PacemakeR), from
which it installs directly.

## Introduction

Marathons are one of the biggest sport-events in counts of active particpants worldwide. One runs 42.195 km. As challenging as it sounds, having a strategy how to optimally distribute energy over the course of the race can make the journey more enjoyable. *This package is an attempt to provide an optimal energy conserving pacing strategy.*

The target group for this package is the ambitious top 15% of amateur runners who aim to run a marathon in under 3 hours or less. In fact the target group is me as I aim to run Sub 2:30 end of this year.

#### A few running terms, briefly

For readers who have never run a marathon, four terms are worth defining up
front, since everything below is built on them:

- **Splits** are recorded by races as the the elapsed
  time between fixed checkpoints (timing mats) along the course, e.g. every 5 km, the half (21.0975 km), and the finish (42.195 km).
- **Pace** is the speed expressed as time per kilometer, written `mm:ss` (e.g. a pace
  of `04:30` means four minutes thirty seconds per km). Lower pace = faster.
  Pace is just the inverse of speed, scaled to the units runners think in.
- **Relative Pace** is the pace of a split relative to some split normalised to 1. With 0.98 being 2% slower and 1.02 being 2% faster.
As default the normalised pace = 1 is at 25 km, as it's a good proxy for potential average pace without being biased of either early enthusiasm or late race decline.
- **Pacemaker** is the person that dictates the pace in a race, hence the name of this package.

### The optimal strategy
For any marathon‑running economist (like me), the optimal pacing strategy naturally becomes an intertemporal optimization problem: start too fast and you collapse under fatigue constraints also called "hitting the wall"; start too slow and you under‑use your physiological budget. In theory the optimal strategy is to run a marathon with even splits, always running the same speed, however in reality this rarely works and leads to disastrous blow ups. The alternative is to run a *negative split*.
That is, saving energy in the beginning to cover the second half of the
race faster than the first. It is widely regarded as the hallmark of mastery,
showcasing discipline and precision, and the latest world records have been set
this way. It reflects the delicate balance between conserving energy early and
maximising output late.
This package assumes that running a negative split is the solution to the described problem above.

Whether negative splitting should be the goal is a matter of
debate. The negative-split approach is essentially risk minimization: it lets
you start conservatively and minimise blow-up risk, assuming you can set a
realistic goal. On a good day you negative split and reach the target; on a bad
day you do not reach the target time but still get close to *even splitting*. Since marathon performance comes down to months and years of training, chance plays only a minor role, and
there is rarely a "surprise" performance that is fast outside a fairly predictable range.

This package analyses marathon split data through that lens. It evaluates a
runner's pacing relative to the rest of the field, which filters out fixed effect
factors such as weather, temperature, or course features like the elevation profile. This comparative approach gives a cleaner read on pacing and describes the split-wise pacing strategy the data actually rewards.

While the pacing progression of an individual runner is somewhat endogenous, my experience and expertise give me the confidence to say that the results in the following case study are sensible and can be utilized in the next marathon one runs.


# Case Study
The following case study introduces the package functions and explains its logic, as well as analyse the 2026 London Marathon. I genuinely hope this package fits somewhat to what is expected from this course!

### Installation

You can install the development version of PacemakeR from
[GitHub](https://github.com/Runandecon/PacemakeR) with:

``` r
# install.packages("devtools")
devtools::install_github("Runandecon/PacemakeR")
library(PacemakeR)
```

### The data

The package (currently) has one example dataset of the `London_Marathon_2026`, the split-level results
of the fastest 6000 **amateur men** in the 2026 London Marathon, scraped from the
[official results website](https://results.tcslondonmarathon.com/2026/?pid=start&pidp=start).  I am currently working integrating my scraping method as a function to this package, but I did not manage to make it work before the deadline. The idea of the package is that one can repeat the analysis for any Marathon-data available, or evaluate ones own performance.

```{r download, echo=FALSE, message=FALSE, warning=FALSE}
library(devtools)
devtools::install_github("Runandecon/PacemakeR")
library(PacemakeR)
```

`PacemakeR` expects the data to be structured in this way before starting the analysis. The columns are `Name`, `Nationality`, `Bib` (the runner's id number), `Distance` (the split
distance in km), and `Time` (elapsed time at that split, as `hh:mm:ss`).
```{r data}
data("London_Marathon_2026")
head(London_Marathon_2026)
```

Marathon results are usually clean with some messy exceptions:
missing splits, runners who dropped out,
times recorded as blanks or `DNF`. Before any analysis is possible the data has to be reshaped from this long format into one tidy pacing table per runner and cleaned of those invalid entries. The package does this with a **tidyverse** (`dplyr` / `tibble`) pipeline grouping by runner, ordering splits, and differencing successive times into segment paces. This is the data-manipulation backbone everything downstream relies on.

## The functions

The functions build on each other: describe the field, distil its pacing
strategy, and price the payoff of running it well. Each one also leans on a
particular tool from the course, introduced where it does its work.

### `marathon_summary()` — describe the field

`marathon_summary()` offers descriptive statistics. The function reduces the field to a five-number summary (mean, median,
min, max, standard deviation) of either finishing times or any intermediate
split. It can report this in **time** (`hh:mm:ss`) or in **pace** (`mm:ss` per
km), and can draw the distribution.

The distribution plot is the package's first use of **ggplot2**, which
handles all visualisation here. On top of the histogram it can overlay vertical references using `markers`; the two below mark the 02:30:00 hours and a 03:00:00 hours finish time, which the function converts automatically from a finish time into the pace it implies, so a target on the time scale lines up correctly on the pace axis.

```{r summary-table}
marathon_summary(London_Marathon_2026)
```

The first plot shows the distribution of the runners at 5 km. You can see peaks at the paces necessary for the red marked goals of 02:30:00 and 03:00:00. This shows the clear intend that most runners aim to run below these target times. The plot suggests to be normally distributed.

```{r pacing-distribution-5K}
marathon_summary(London_Marathon_2026, distance = 5, pace = TRUE, plot = TRUE, markers = c("02:30:00", "03:00:00"))
```


The second plot shows the finishing time distribution. It is clear that the distribution of the very same runners is now skewed to the left, as many runners overshoot in the beginning part of the race.

```{r pacing-distribution-Finish}
marathon_summary(London_Marathon_2026, plot = TRUE, pace = TRUE, markers = c("02:30:00", "03:00:00"))
```

### `plot_pace_splits()`
One can also view the race of any individual using `plot_pace_splits()`.
```{r pacing-distribution-individual}
plot_pace_splits(London_Marathon_2026, id = 1000)
```
## The optimal strategy

### `pacemaker()` — distil the strategy

`london <- pacemaker()` is the analytical core. For every runner it builds a
*relative-pace curve*: their pace at each split, normalised so that a chosen
reference split equals 1. Values above 1 mean "faster than the reference," below 1 "slower." The aggregate curve is the field's average index path; the band is its empirical spread.

Normalising this way is what makes a fast and a slow
runner comparable on the same axis. This strips out absolute speed and leaves
only the *shape* of the effort, i.e. where each runner spent or conserved
energy. The function then aggregates these curves across the whole field, with a quantile band for the spread, and separately averages the runners who ran a
negative split, overlaying them for comparison. Given the nature of the runners in the data (ambitious amateurs) assume they are all comparable. In fact, the data suggests so when I tested the analysis in sub-groups.
 

This is where the package's use of **S3 classes** comes in. Rather than handing back a bare list of numbers, `pacemaker()` returns an object of a custom class, `pacemaker`, that bundles the curve, the cumulative-advantage data, and metadata about the fit. Defining a class lets the package give the object its own `print` and `plot` methods, so it behaves like a first-class R object: typing its name dispatches to `print.pacemaker`, which lays out the pacing progression as a table, and calling `plot()` dispatches to `plot.pacemaker`, which draws the
curve. The single object is the whole analysis. Everything you might want to
see is computed once and read off it on demand.

```{r curve}
london <- pacemaker(London_Marathon_2026, relative = 25, agg_fun = "median")
class(london)
```

### `plot(london)` show the median
The blue line is the field's median relative pace at each split, normalised so the 25 km reference equals 1; the grey band is the 5th–95th percentile spread across runners, and the dashed purple line is the average curve of the negative-split runners. Values above 1 are faster than reference pace, below 1 slower. The typical runner starts slightly quick, holds steady through halfway, then fades after 25 km (of course that is by design, but 25K is usually the last split before many "hit the wall"). the band widening toward the finish shows how much that late fade varies between runners. The negative-split group (purple) does the opposite: it stays just below the field early and pulls ahead late, exactly the pattern the package argues for. More importantly. Notably, the early median slowdown is faster then the later median decline between of the other runners, thus in the long run more more efficient.

```{r curve-plot}
plot(london)
```

### `plot(london, "gain")` — price the payoff

Because the analysis lives in one `pacemaker` object, pricing the strategy is
just another view of it rather than a separate function. The object also stores
the *gain*: the cumulative advantage a negative-split runner builds over the
field across the race. Asking its `plot` method for the `"gain"` view integrates
the gap between the two curves along the course and reports it either as distance
(metres ahead (similar for any pace) or, given a reference pace, as time saved — the same S3 `plot`
method, switched to a different output.
The plot shows the cumulative integral of the per-segment
difference between the two paths. The running total of how far ahead the
negative-split strategy puts you, expressed in units a runner cares about.
Many assumptions are made to make this outcome pessimistic to add robustness. Following the negative split progression one safes 300m worth of distance. Or 0.3 times the pace at 25 km, which in the realm of this analysis is 60 to 100 seconds. This results in a significant time save, that one can hopefully utilize in the next competition.

```{r gain}
plot(london, "gain")                                  
```

```{r print}
print(london)                               
```

## Course tools at a glance

The tools above were each introduced where they do their work: version control
and the package skeleton frame the whole project; the **tidyverse** pipeline
reshapes and cleans the raw splits; **ggplot2** carries every visualisation; and
a custom **S3 class** turns the central `pacemaker()` result into an object that
prints and plots itself. The package is tracked on GitHub, documented with
`roxygen2`, and validated by a `testthat` suite covering the helpers, the core
computation, and the S3 methods, so `devtools::check()` passes from a clean
environment.
